---
title: "1 Preparing your structure-organism pairs library"
author: "Adriano Rutz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1 Preparing your structure-organism pairs library}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes how to prepare a structure-organism pairs library in order to

* Annotate or complement the annotation of your feature. *(optional)*
* **Weight your initial annotations**.

## LOTUS

As we developed also [LOTUS](https://lotusnprod.github.io/lotus-manuscript) with 
**T**axonomically 
**I**nformed 
**M**etabolite 
**A**nnotation in mind, 
we provide it here as a starting point for your structure-organism pairs library. 
You are free to complement it with your own structure-organism pairs or to filter it according to your needs. We also provide some hints on how to do it later in this vignette.


To download it LOTUS structure-organism pairs locally, just run:

```{r}
library("timaR")
setwd(dir = "../") ## adapt it in order to be in timaR root directory
source(file = "inst/scripts/get_lotus.R")
```
