---
title: "2 Formatting all inputs"
author: "Adriano Rutz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2 Formatting all inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = '..')
```

This vignette describes how to format all your files.

## Structural annotations of your features

For the moment, we support annotations coming from 3 different annotation tools: [GNPS](https://gnps.ucsd.edu), [SIRIUS](https://bio.informatik.uni-jena.de/software/sirius) and our *in lab* developed tool, **[ISDB](https://github.com/mandelbrot-project/spectral_lib_matcher)**.
We hope to soon implement SIRIUS support.

In order to perform MS2 annotation based on an **I**n **S**ilico **D**ata**B**ase, please follow the following repository: [spectral_lib_matcher](https://github.com/mandelbrot-project/spectral_lib_matcher)

For Sirius and GNPS, please follow their own documentation.

### Get an example

#### ISDB

As [spectral_lib_matcher](https://github.com/mandelbrot-project/spectral_lib_matcher) is only in python, for the moment, we provide an example output coming from the new ISDB.
We are working on a better integration. 
As this comes by default, you do not need to modify any parameters

```{r}
library("timaR")
```

```{r}
source(file = "inst/scripts/get_example_isdb.R")
```

#### SIRIUS
As, given its size, it is not easy to share a complete SIRIUS job, we prepared an example containing only summary files.
We also generated the npc classes as in https://github.com/kaibioinfo/canopus_treemap.
They have to be stored in the same folder under the name 'canopus_npc_summary.csv'.

You can get the example running:
```{r}
source(file = "inst/scripts/get_example_sirius.R")
```

#### GNPS

We also provide an example GNPS job id, which is: `96fa7c88200e4a03bee4644e581e3fb0`

### Formatting

Before running the corresponding code, do not forget to modify `config/params/prepare_*yourAnnotationTool*.yaml`.

Depending on the annotation tool you used, you can format its results using: 

```{r}
source(file = "inst/scripts/prepare_gnps.R")
```

or 

```{r}
source(file = "inst/scripts/prepare_isdb.R")
```

or 

```{r}
source(file = "inst/scripts/prepare_sirius.R")
```

You now have your annotations well prepared and can keep on formatting the rest of your metadata

## Chemical class annotation of your features

Within our workflow, we offer a new way to attribute chemical classes to your features.
It is analog to [Network Annotation Propagation](https://ccms-ucsd.github.io/GNPSDocumentation/nap/), but uses the edges of your network instead of the clusters.
This makes more sense in our view, as also recently illustrated by [CANOPUS](https://doi.org/10.1038/s41587-020-0740-8).

All steps can take both manual inputs or GNPS metadata directly from your GNPS job ID.

We are currently also working on CANOPUS integration for chemical class annotation but this implies way heavier computations and we want to offer our users a **fast** solution.

### Formatting

Before running the corresponding code, do not forget to modify `config/params/prepare_edges.yaml`, `config/params/prepare_features_components.yaml`, and `config/params/prepare_features_classification.yaml` accordingly.

#### Case when no network available
If no network was generated, simply use the `fake_edges` and `fake_features_components` steps instead of the `prepare_edges` and `prepare_features_components` steps. This will mimic a false network that allows to proceed to the next steps.

However, this will not help chemical weighting.

```{r}
source(file = "inst/scripts/prepare_edges.R")
source(file = "inst/scripts/prepare_features_components.R")
source(file = "inst/scripts/prepare_features_classification.R")
```

Your features are now not only informed with structural information but also, chemical class information.
The latter might be corresponding **or not** to the chemical class of your annotated structure, depending on the consistency of your annotations.

## Biological source annotation

This step allows you to attribute biological source information to your features.
If all your features come from a single extract, it will attribute the biological source of your extract to all your features.
If you have multiple extracts aligned, it will take the *n* (according to your parameters) highest intensities of your aligned feature table and attribute the biological source of corresponding extracts.
It can take both manual inputs or GNPS metadata directly from your GNPS job ID.

Before running the corresponding code, do not forget to modify `config/params/prepare_taxa.yaml`.

```{r}
source(file = "inst/scripts/prepare_taxa.R")
```

We now recommend you to read the [next vignette](https://taxonomicallyinformedannotation.github.io/tima-r/articles/3-processing.html).
