---
title: "2 Inputs: Getting everything you need"
author: "Adriano Rutz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2 Inputs: Getting everything you need}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = "..")
```

This vignette describes how to prepare:

* All parameters you need
* All inputs you need

```{r include = FALSE}
paths <- yaml::read_yaml(
  file = "paths.yaml",
  handlers = list(
    seq = function(x) {
      purrr::list_flatten(x)
    }
  )
)

paths$tests$mode <- "yes"
paths$urls$examples$spectra <- paths$urls$examples$spectra_mini

yaml::write_yaml(x = paths, file = "paths.yaml")
```

## Parameters

As mentioned previously, almost all steps require parameters.
Some pre-defined parameters are given in `config/default` directory.
Those will help you understanding how the workflow works so you can adapt them later to your own data.

So first, let's have a look at the structure of the parameters:

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("config_default"), targets_only = TRUE, degree_from = 8)
```

### Copy

Now, you want to copy them and adapt them to your own files to get something like:

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("config"), targets_only = TRUE, degree_from = 8)
```

### Prepare

To do so, simply:

```{r}
source(file = "inst/scripts/prepare_config.R")
```

And now, you should get the following:

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("config"), targets_only = TRUE, degree_from = 8)
```

See how your targets changed?
The values you adapted in `config/prepare_config.yaml` were automatically propagated to all subfiles in `config/user`.

And everything is stored in case you re-perform it, the beauty of `{targets}`!

```{r}
source(file = "inst/scripts/prepare_config.R")
```

You can easily fine tune all parameters in each file and only the steps impacted by these modifications will run!

### Adapt

Your final prepared parameters look now like:

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("params"), targets_only = TRUE, degree_from = 8)
```

## Inputs

### Dictionaries

The package already comes with some default dictionaries.
These dictionaries are needed among the pipeline and it is recommended **NOT** to touch them.
If you are confident in doing so, feel free.

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("dic"), targets_only = TRUE, degree_from = 8)
```

### Your own file(s)

Basically, the most important file you need to provide is an MGF file containing the spectra you want to annotate, as you can see in the architecture:

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("input"), targets_only = TRUE, degree_from = 8)
```

To get an example of an MGF file (corresponding to the spectra to annotate, not to to the library to query), just run:

```{r}
source(file = "inst/scripts/get_example_spectra.R")
```

### Libraries

#### Structure-Organism Pairs

##### LOTUS

As we developed [LOTUS](https://lotusnprod.github.io/lotus-manuscript)^[For more informations, see <https://doi.org/10.7554/eLife.70780>] with
**T**axonomically
**I**nformed
**M**etabolite
**A**nnotation in mind,
we provide it here as a starting point for your structure-organism pairs library.

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("input"), targets_only = TRUE, degree_from = 8)
```

You can easily get it by running: 

```{r}
source(file = "inst/scripts/get_lotus.R")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = matches("library_sop_lotus$"), targets_only = TRUE, degree_from = 8)
```

You then need to format it as expected by the pipeline, to do so:

```{r}
source(file = "inst/scripts/prepare_lotus.R")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = matches("library_sop_lotus"), targets_only = TRUE, degree_from = 8)
```

As you can see, the target seems outdated.
In reality, we force it to search if a new version of LOTUS exists each time.
If a newer version exists, it will fetch it and re-run needed steps accordingly.

##### HMDB

If you want, you can also complement LOTUS pairs with the ones coming from HMDB (not running by default as quite long):

```{r eval=FALSE, include=TRUE}
source(file = "inst/scripts/get_hmdb.R")
```

As previously, you then need to prepare the library:

```{r eval=FALSE, include=TRUE}
source(file = "inst/scripts/prepare_hmdb.R")
```

For these first steps, you do not need to change any parameters as they are implemented *by default*.

##### Other libraries

As we want our tool to be flexible, you can also add your own library to LOTUS.
You just need to format it in order to be compatible.
As example, we prepared some ways too format closed, *in house* libraries.
If you need help formatting your library or would like to share it with us for it to be implemented, feel free to contact us.

Before running the corresponding code, do not forget to modify `config/user/prepare_closed.yaml`

```{r}
source(file = "inst/scripts/prepare_closed.R")
```

##### Merging

Once you have all your sub-libraries prepared, you are ready to merge them in a single file that will be used for the next steps.

Before running the corresponding code, do not forget to modify `config/user/prepare_libraries.yaml`.

At this step, if you want to, you can restrict your library to specific taxa only.
We do not advise doing so, but we offer you the possibility to do it.

```{r}
source(file = "inst/scripts/prepare_libraries.R")
```

You now should have a nice custom structure-organism pairs library for the next steps!

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("library_sop"), targets_only = TRUE, degree_from = 8)
```

#### Adducts

The next library you need is an adducts library!
As a main limitation of current annotation tools is adducts detection and coverage, an adducts library can also be generated in order to perform MS1 annotation later on.

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("library_adducts"), targets_only = TRUE, degree_from = 8)
```

As you can see, it depends on the previously built library.

Before running the corresponding code, do not forget to modify `config/user/prepare_adducts.yaml`.

```{r}
source(file = "inst/scripts/prepare_adducts.R")
```

#### Spectra

Finally, you need a spectral library to perform MS2-based annotation.

##### Experimental

You can of course use your own experimental spectral library to perform MS2 annotation.
We currently only support spectral libraries in MGF format.

We are currently working to get all MONA and GNPS experimental spectra easily available.

However, programmatic download of MONA is currently not supported, and GNPS spectra require some pre-processing.

So for now, either use your GNPS job ID and download MONA manually, helpers to use them later on are already available.

##### *In silico*

As the availability of experimental spectra is limited, we can take advantage of *in silico* generated spectra.

#### LOTUS

We generated an *in silico* spectral library of the structures found in LOTUS using [CFM4](https://doi.org/10.1021/acs.analchem.1c01465).
For more info, see <https://doi.org/10.5281/zenodo.5607185>.

In order to get it:

```{r}
source(file = "inst/scripts/get_isdb_lotus.R")
```

And as previously, then prepare it:

```{r}
source(file = "inst/scripts/prepare_isdb_lotus.R")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("library_spectra"), targets_only = TRUE, degree_from = 8)
```

As you can see, both MS polarities are available!

Again, you can also complement with the in silico spectra from HMDB (not running by default as quite long):

#### HMDB

```{r eval=FALSE, include=TRUE}
source(file = "inst/scripts/get_isdb_hmdb.R")
```

You are finally ready for the next step, with all dictionaries, inputs, and libraries you need!

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
targets::tar_visnetwork(names = starts_with("library"), targets_only = TRUE, degree_from = 8)
```

We now recommend you to read the [next vignette](https://taxonomicallyinformedannotation.github.io/tima-r/articles/III-formatting.html).
