---
title: "1 Preparing your structure-organism pairs library"
author: "Adriano Rutz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1 Preparing your structure-organism pairs library}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = '..')
```

This vignette describes how to get and prepare a structure-organism pairs library in order to

* Annotate or complement the annotation of your feature. *(optional)*
* **Weight your initial annotations**.

## LOTUS

As we developed [LOTUS](https://lotusnprod.github.io/lotus-manuscript)^[For more informations, see <https://lotus.nprod.net/>] with 
**T**axonomically 
**I**nformed 
**M**etabolite 
**A**nnotation in mind, 
we provide it here as a starting point for your structure-organism pairs library. 
You are free to complement it with your own structure-organism pairs or to filter it according to your needs. We also provide some hints on how to do it later in this vignette.

### download

To download it LOTUS structure-organism pairs locally, just run:

```{r}
library("timaR")
```

```{r}
source(file = "inst/scripts/get_lotus.R")
```

### preparation

Then, **LOTUS** structure-organism pairs can be prepared to the right format for further processing:
```{r}
source(file = "inst/scripts/prepare_lotus.R")
```

For these two first steps, you do not need to change any parameters as they are implemented *by default*. 

## Other libraries

As we want our tool to be flexible, you can also add your own library to LOTUS.
You just need to format it in order to be compatible.
As example, we prepared some ways too format closed, *in house* libraries.
If you need help formatting your library or would like to share it with us for it to be implemented, feel free to contact us.

Before running the corresponding code, do not forget to modify `config/params/prepare_closed.yaml`

```{r}
source(file = "inst/scripts/prepare_closed.R")
```

## Merging

Once you have all your sub-libraries prepared, you are ready to merge them in a single file that will be used for the next steps.

Before running the corresponding code, do not forget to modify `config/params/prepare_library.yaml`.

At this step, if you want to, you can restrict your library to specific taxa only.
We do not advise doing so, but we offer you the possibility to do it.

```{r}
source(file = "inst/scripts/prepare_library.R")
```

## Adducts library

As a main limitation of current annotation tools is adducts detection and coverage, an adducts library can also be generated in order to perform MS1 annotation later on.

Before running the corresponding code, do not forget to modify `config/params/prepare_adducts.yaml`.

```{r}
source(file = "inst/scripts/prepare_adducts.R")
```

We now recommend you to read the [next vignette](https://taxonomicallyinformedannotation.github.io/tima-r/articles/2-formatting.html).

