---
title: "4 Performing Taxonomically Informed Metabolite Annotation"
author: "Adriano Rutz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4 Performing Taxonomically Informed Metabolite Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = "..")
```

This vignette describes how to perform **T**axonomically **I**nformed **M**etabolite **A**nnotation.
If you followed all previous steps successfully, this should be a piece of cake, you deserve it!

```{r, message=FALSE, warning=FALSE}
library("timaR")
```

Before running the corresponding code, do not forget to modify `params/user/weight_annotations.yaml`.

```{r, message=FALSE}
source(file = "inst/scripts/weight_annotations.R")
```

The final exported file is formatted in order to be easily imported in [Cytoscape](https://cytoscape.org/) to further explore your data!

We hope you enjoyed using **TIMA** and are pleased to hear from you!

For any remark or suggestion, please  [fill an issue](https://github.com/taxonomicallyinformedannotation/tima-r/issues) or feel free to contact us directly.
