---
title: 2 Preparing inputs
author:
  name: Adriano Rutz
  orcid: 0000-0003-0443-9902
citation:
  url: https://taxonomicallyinformedannotation.github.io/tima/vignettes/articles/II-preparing.html
comments:
  giscus:
    repo: taxonomicallyinformedannotation/tima
  hypothesis:
    showHighlights: always
creative_commons: CC BY
date: today
format:
  html:
    smooth-scroll: true
google-scholar: true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
lang: en
license: CC BY
opengraph:
  image:
    src: https://github.com/taxonomicallyinformedannotation/tima/blob/main/man/figures/logo.svg
    alt: Taxonomically Informed Metabolite Annotation
vignette: >
  %\VignetteIndexEntry{2 Preparing inputs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: tima
#| results: "hide"
#| message: false
#| warning: false
#| include: false

library("tima")
if (dir.exists("../../inst")) {
  knitr::opts_knit$set(root.dir = "../../inst")
}
logger <- lgr::lgr
logger$set_threshold(level = 400)
```

This vignette describes the main steps of the annotation process.

## Structural annotations of your features

For the moment, we support 3 different types of annotations:

<!-- * [GNPS](https://gnps.ucsd.edu) -->

- Internal MS^1^ exact mass-based library search
- Internal MS^2^ library search (experimental and in silico)
- [SIRIUS](https://bio.informatik.uni-jena.de/software/sirius)

If needed, you can get an example of what your minimal feature table should look
like by running:

```{r}
#| label: get-features
#| eval: false
#| include: true

tima::get_example_files(example = "features")
```

Then, you can prepare your features for the next steps:

```{r}
#| label: prepare-features
#| eval: false
#| include: true

tima::prepare_features_tables()
```

### MS^1^-based

These annotations are of the lowest possible quality. However, they allow to
annotate unusual adducts, in-source fragments thanks to different small tricks
implemented. Try to really restrict the adduct list and structure-organism pairs
you want to consider as possibilities explode rapidly.

```{r}
#| label: annotate-masses-1
#| eval: false
#| include: true

tima::annotate_masses()
```

#### How MS1 adduct attribution works

`annotate_masses()` uses two levels:

- **Primary (strict graph):** each `feature_id` gets one primary ion species.
  Edges are kept only when both nodes are strongly supported, and consistent
  with the same neutral mass `M`.
- **Secondary (coverage):** additional non-conflicting ion-species hypotheses
  are kept as secondary.

For unresolved nodes in a supported component, TIMA also tests direct forms from
the component mass `M` (for example `[M +/- loss + carrier]`) and keeps valid
ones as weak support.

Main output fields to inspect:

- `candidate_adduct`
- `candidate_adduct_origin` (`supported`, `supported_weak`, `baseline`)
- `candidate_annotation_level` (`primary`, `secondary`)
- `candidate_evidence_tier` (`supported_strong`, `supported_weak`, `baseline`)

### Spectral

```{r}
#| label: annotate-spectra-1
#| eval: false
#| include: true

tima::annotate_spectra()
tima::prepare_annotations_spectra()
```

There are multiple similarities available, including `spectral entropy` from
<https://doi.org/10.1038/s41592-021-01331-z> for matching.

#### GNPS

Optional

```{r}
#| label: prepare-gnps-1
#| eval: false
#| include: true

tima::prepare_annotations_gnps()
```

### Fingerprint-based

#### Sirius

As SIRIUS jobs are long to perform, we provide example SIRIUS workspaces (both
SIRIUS 5 and 6). Note that spectral matches from SIRIUS are not supported for
now. They have been generated on the 20 first lines of the example MGF with the
following command:

```{bash}
#| label: run-sirius
#| eval: false
#| include: true

# this is run on SIRIUS 6
sirius \
--noCite \
--input=data/source/example_spectra_mini.mgf \
--output=data/interim/annotations/example_sirius.sirius/ \
--maxmz=800 \
config \
--AlgorithmProfile=orbitrap \
--StructureSearchDB=BIO \
--Timeout.secondsPerTree=10 \
--Timeout.secondsPerInstance=10 \
formulas \
zodiac \
fingerprints \
classes \
structures \
denovo-structures \
summaries \
--chemvista \
--feature-quality-summary \
--full-summary

# this is run on SIRIUS 5
sirius \
--noCite \
--input data/source/example_spectra_mini.mgf \
--output data/interim/annotations/example_sirius/ \
--maxmz 800 \
config \
--AlgorithmProfile orbitrap \
--StructureSearchDB BIO \
--Timeout.secondsPerTree 10 \
--Timeout.secondsPerInstance 10 \
formula \
zodiac \
fingerprint \
structure \
compound-classes \
write-summaries \
--full-summary
```

These parameters were not optimized and were only used to give an example
output. If you are using the cli, do not forget to generate the summaries with
the `--full-summary` option, or if you use the gui, generate them by clicking
the corresponding icon. You can get an example running:

```{r}
#| label: get-sirius
#| eval: false
#| include: true

getFromNamespace("get_example_sirius", "tima")()
```

The sirius workspace should ideally have `yourPattern_sirius` as name and be
placed in `data/interim/annotations` (else it will not be found by default
except you provide the right path).

```{r}
#| label: prepare-sirius
#| eval: false
#| include: true

tima::prepare_annotations_sirius()
```

If you want to know how we attempt to combine the CSI score with other ones, see
[R/transform_score_sirius_csi.R](https://github.com/taxonomicallyinformedannotation/tima/blob/main/R/transform_score_sirius_csi.R)
Note that starting from SIRIUS 6, the *approx* confidence score is the one
considered, and not the exact one.

```{r}
#| label: annotations-all
#| echo: false
#| message: false
#| warning: false
#| out-width: "100%"

targets::tar_visnetwork(
  names = starts_with("ann_s"),
  exclude = targets::contains(
    c(
      "benchmark",
      "par_",
      "lib_",
      "paths",
      "test"
    )
  ),
  targets_only = TRUE,
  degree_from = 8
)
```

Annotations are now prepared and can be used for further processing. Your
features are not only informed with structural information but also, chemical
class information. The latter might be corresponding **or not** to the chemical
class of your annotated structure, depending on the consistency of your
annotations.

## Chemical class annotation of your features

Within our workflow, we offer a new way to attribute chemical classes to your
features. It is analog to [Network Annotation
Propagation](https://ccms-ucsd.github.io/GNPSDocumentation/nap/), but uses the
edges of your network instead of the clusters. This makes more sense in our
view, as also recently illustrated by
[CANOPUS](https://doi.org/10.1038/s41587-020-0740-8).

<!-- All steps can take both manual inputs or GNPS metadata directly from your GNPS job ID. -->

We are currently also working on CANOPUS integration for chemical class
annotation but this implies way heavier computations and we want to offer our
users a **fast** solution.

### Generating a network

A network is generated during the process. The edges are created based on the
`spectral entropy similarity` calculated between your spectra (see
<https://doi.org/10.1038/s41592-021-01331-z>).

```{r}
#| label: create-edges
#| eval: false
#| include: true

tima::create_edges_spectra()
```

Prepare the edges:

```{r}
#| label: prepare-edges
#| eval: false
#| include: true

tima::prepare_features_edges()
```

Create and prepare the components:

```{r}
#| label: create-components
#| eval: false
#| include: true

tima::create_components()
```

```{r}
#| label: prepare-components
#| eval: false
#| include: true

tima::prepare_features_components()
```

## Biological source annotation

This step allows you to attribute biological source information to your
features. If all your features come from a single extract, it will attribute the
biological source of your extract to all your features. If you have multiple
extracts aligned, it will take the *n* (according to your parameters) highest
intensities of your aligned feature table and attribute the biological source of
corresponding extracts.
<!-- It can take both manual inputs or GNPS metadata directly from your GNPS job ID. -->

```{r}
#| label: prepare-taxa
#| eval: false
#| include: true

tima::prepare_taxa()
```

## Filter annotations (based on retention time)

This step allows you to filter out the annotation of all the tools used, based
on your own internal (experimental or predicted) retention times library. It is
optional. If you do not have one, it will simply group the annotations of all
tools.

```{r}
#| label: filter-annotations
#| eval: false
#| include: true

tima::filter_annotations()
```

You are almost there! See already all the steps accomplished!

```{r}
#| label: features-all
#| echo: false
#| message: false
#| warning: false
#| out-width: "100%"

targets::tar_visnetwork(
  names = tidyselect::matches("^ann_fil"),
  exclude = targets::contains(
    c(
      "benchmark",
      "par_",
      "lib_",
      "paths",
      "test"
    )
  ),
  targets_only = TRUE,
  degree_from = 8
)
```

We now recommend you to read the [next
vignette](https://taxonomicallyinformedannotation.github.io/tima/vignettes/articles/III-processing.html).
