---
title: "3 Performing Taxonomically Informed Metabolite Annotation"
author: "Adriano Rutz"
creative_commons: CC BY-SA
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
opengraph:
    image:
      src: https://github.com/taxonomicallyinformedannotation/tima/blob/main/man/figures/logo.svg
      alt: "Taxonomically Informed Metabolite Annotation"
    twitter:
      creator: "@adafede"
vignette: >
  %\VignetteIndexEntry{3 Performing Taxonomically Informed Metabolite Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = system.file(package = "tima"))
```

```{r tima, results="hide", message=FALSE, warning=FALSE, include = FALSE}
library("tima")
```

This vignette describes how **T**axonomically **I**nformed **M**etabolite **A**nnotation is performed.
If you followed all previous steps successfully, this should be a piece of cake, you deserve it!

```{r weight-all, message=FALSE}
targets::tar_make(names = tidyselect::matches("ann_pre$"))
```

The final exported file is formatted in order to be easily imported in [Cytoscape](https://cytoscape.org/) to further explore your data!

We hope you enjoyed using **TIMA** and are pleased to hear from you!

For any remark or suggestion, please  [fill an issue](https://github.com/taxonomicallyinformedannotation/tima/issues) or feel free to contact us directly.
