---
title: 3 Performing Taxonomically Informed Metabolite Annotation
author: 
  - name: Adriano Rutz
    orcid: 0000-0003-0443-9902
comments:
  hypothesis:
    showHighlights: always
creative_commons: CC BY-SA
date: "`r Sys.Date()`"
format:
  html:
    # code-links:
    #  - repo: https://github.com/taxonomicallyinformedannotation/tima
    smooth-scroll: true
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
  opts_knit:
    root.dir: ../../inst
opengraph:
    image:
      src: https://github.com/taxonomicallyinformedannotation/tima/blob/main/man/figures/logo.svg
      alt: Taxonomically Informed Metabolite Annotation
vignette: >
  %\VignetteIndexEntry{3 Performing Taxonomically Informed Metabolite Annotation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r tima, results="hide", message=FALSE, warning=FALSE, include = FALSE}
library("tima")
```

This vignette describes how **T**axonomically **I**nformed **M**etabolite **A**nnotation is performed.
If you followed all previous steps successfully, this should be a piece of cake, you deserve it!

```{r weight-all, message=FALSE}
targets::tar_make(names = tidyselect::matches("ann_pre$"))
```

The final exported file is formatted in order to be easily imported in [Cytoscape](https://cytoscape.org/) to further explore your data!

We hope you enjoyed using **TIMA** and are pleased to hear from you!

For any remark or suggestion, please  [fill an issue](https://github.com/taxonomicallyinformedannotation/tima/issues) or feel free to contact us directly.
